---
title: "00_wrangle_metadata"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# Rscript -e "rmarkdown::render('src/00_wrangle_metadata.Rmd', output_dir = 'out/metadata/')" # nolint: line_length_linter.
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)

# dirs
data_dir <- "/nfs/cancer_ref01/nst_links/live/2731"
dir.create("out/merged_normal_bams/", showWarnings = F)
dir.create("out/nanoseq/", showWarnings = F)

# libraries
library(magrittr)

# function: clean up metadata
cleanup_metadata <- function(df) {
  if ("donor_sex" %in% colnames(df)) {
    df <-
      df %>%
      dplyr::mutate(
        donor_sex = dplyr::case_when(
          donor_sex %in% c("male", "Male") ~ "M",
          donor_sex %in% c("female", "Female") ~ "F",
          donor_sex == "Unknown" ~ NA_character_,
          TRUE ~ donor_sex))
  }
  df %>%
    # re-check col types
    readr::type_convert() %>%
    dplyr::distinct()
}

# dirs
dir.create("out/metadata/", showWarnings = FALSE)
```

# Introduction

This script wrangles the metadata from the project 
**NanoSeq - (CP) Normal and cancer-predisposed kidney (2731)**.

## Handling LCM metadata

There are three levels to the metadata in an LCM experiment that I have formalised for this project:

1. **Donor-level metadata** - This is all of the information that is unique to 
each donor. Each donor is defined by their PDID (e.g. PD12345).

2. **Sample-level metadata** - This is all of the information that is unique to 
each sample. In the case of LCM, a sample is defined as one slide of sections 
from the donor (e.g. PD12345a).

3. **Biopsy-level metadata** - This is all of the information that is unique to 
each biopsy. In the case of LCM, a biopsy is defined as one well of cuts from 
the sample that is sequenced (e.g. PD12345a_ds0001).

In the output files, information pertaining to each of these levels will be 
prefixed with `donor_`, `sample_`, or `biopsy_`.

# Inputs (in `data/`)

When I began this project, I was given various files describing the dataset that I needed to harmonise into a consistent, clean, centralised metadata file in order to run the analyses.
I describe these files here.

## `LCM_log_byStructure_kidney.xslx`

Sent by cp19@sanger.ac.uk, 19.03.2024

```{r class.source = 'fold-hide'}
readxl::read_xlsx("data/LCM_log_byStructure_kidney.xlsx") %>%
  knitr::kable()
```

This is an overview of the number of samples cut per patient per histological structure. 

```{r}
lcm_log <-
    readxl::read_xlsx("data/LCM_log_byStructure_kidney.xlsx") %>%
    dplyr::filter(Sample != "Total") %>%
    dplyr::transmute(
        sample_id = Sample,
        sample_tumour_status = `Tumor?` > 0,
        donor_id = sample_id %>% stringr::str_sub(end = -2),
        donor_predisposition = dplyr::case_when(
            `Predisposition?` == "NA" ~ NA_character_,
            TRUE ~ `Predisposition?`),
        donor_diagnosis = Diagnosis) %>%
    # clean up metadata
    cleanup_metadata()

# view
lcm_log %>%
  knitr::kable()
```

## `patientManifest.txt`

Sent by cp19@sanger.ac.uk, 19.03.2024

```{r class.source = 'fold-hide'}
readr::read_tsv("data/patientManifest.txt") %>%
  knitr::kable()
```

This contains details on the donors. This is the main trusted source of donor 
metadata, as the decode files have several inconsistencies due to copying and 
input errors.

```{r}
patient_manifest <-
  readr::read_tsv("data/patientManifest.txt") %>%
  dplyr::transmute(
    donor_id = Sanger,
    donor_predisposition = Syndrome,
    donor_diagnosis = Dx,
    donor_age = Age,
    donor_sex = Sex
  ) %>%
  # clean up metadata
  cleanup_metadata()

# view
patient_manifest %>%
  knitr::kable()
```

## Decode files from Chloe Pacyna

Sent by cp19@sanger.ac.uk, 02.04.2024 & 19.04.2024

The decode files contain experimental information about the LCM cuts, 
connecting them with the donor and sample information. 

4 decode files were created for this project:

```{r}
list.files("data/decode/cp19/", pattern = "xlsx", full.names = TRUE)
```

The following columns were retained to generate the metadata:

```{r class.source = 'fold-hide'}
tibble::tribble(
  ~decode_col,                ~rename,
  "existing_pd_id_for_donor", "sample_id",
  "gender",                   "donor_sex",
  "donor_age_at_diagnosis",   "donor_age",
  "tissue_phenotype",         "biopsy_phenotype",
  "tumour or normal?",        "sample_tumour_status",
  "tissue_histology",         "sample_histology",
  "PD_ID",                    "biopsy_id") %>%
  knitr::kable()
```

```{r}
decodes <-
  list.files("data/decode/cp19/", pattern = "xlsx", full.names = TRUE) %>%
  purrr::set_names(., basename(.)) %>%
  purrr::map(function(file) {
    file %>%
      readxl::read_xlsx(skip = 1) %>%
      dplyr::filter(dplyr::row_number() > 12,
                    !is.na(existing_pd_id_for_donor)) %>%
      dplyr::transmute(
        sample_id = existing_pd_id_for_donor,
        donor_id = stringr::str_sub(sample_id, end = -2),
        donor_sex = gender,
        donor_age = donor_age_at_diagnosis,
        biopsy_phenotype = tissue_phenotype,
        sample_tumour_status = `tumour or normal?`,
        sample_histology = tissue_histology,
        biopsy_id = PD_ID,
        biopsy_bam_file =
          paste0(data_dir, "/", biopsy_id, "/", biopsy_id, ".sample.merged.bam")
      )
  }) %>%
  dplyr::bind_rows(.id = "decode_file") 
```

### Inconsistencies in the decode files

There are several inconsistencies in these files that must be considered.

#### `decode/cp19_kidney2_Sanger_Human_Sample_ProformaV4.0_current_2021_covid.xlsx`

There was an error in the original file which I manually fixed. The `biopsy_id`s 
(e.g. PD51357a_ds0001) were in the `donor_age_at_diagnosis` column, rather than 
the `PD_ID` column.

#### `cp19_kidney4_Sanger_Human_Sample_ProformaV4.0_current_2021_covid.xlsx`

There are two `sample_id`s th`at are mixed up in the `Supplier_Sample_ID` column
in this decode file (PD44966b and PD43948s). For all decode files I source the 
`sample_id` from the `existing_pd_id_for_donor` column, as this column is 
reliably consistent, and then derive the `donor_id` as a substring.

#### Donor metadata errors

##### `PD43948`

PD43948 has metadata from two different donors across 2 of the decode files.

```{r}
decodes %>%
  dplyr::filter(donor_id == "PD43948") %>%
  dplyr::distinct(donor_id, donor_sex, donor_age) %>%
  knitr::kable()
```

According to the `patient_manifest`, PD43948 is a 76-year-old female. 

```{r class.source = 'fold-hide'}
patient_manifest %>%
  dplyr::filter(donor_id == "PD43948") %>%
  knitr::kable()
```

##### `PD47172`

PD47172 also has metadata from two different donors across 2 of the decode files.

```{r}
decodes %>%
  dplyr::filter(donor_id == "PD47172") %>%
  dplyr::distinct(donor_id, donor_sex, donor_age) %>%
  knitr::kable()
```

According to the `patient_manifest`, PD47172 is a 52-year-old male.

```{r class.source = 'fold-hide'}
patient_manifest %>%
  dplyr::filter(donor_id == "PD47172") %>%
  knitr::kable()
```

Because of these issues, the donor information in the patient manifest will
take precedence over the donor information in the decode files.

## Files from Li et al. (2022)

Some of the samples from this project were in the paper "Mapping single-cell 
transcriptomes in the intra-tumoral and associated territories of kidney 
cancer" (Li et al., 2022).

```{r}
li2022_metadata <-
  readr::read_tsv("data/Li_2022/sample_table_from_tjm_20240414.tsv")

# view 
li2022_metadata %>%
  knitr::kable()
```

```{r}
li2022_mmc2 <-
  readxl::read_xlsx("data/Li_2022/mmc2.xlsx", skip = 1)

# view
li2022_mmc2 %>%
  knitr::kable()
```

## Decode files from Thomas Mitchell

Sent by tjm61@medschl.cam.ac.uk, 23.04.2024

Thomas Mitchell also sent metadata and decode files related to this project.

```{r}
list.files("data/decode/tjm", full.names = TRUE)
```

### `decode/tjm/Samples_Tom_24-02-2021_sent.xlsx`

Sent by tjm61@medschl.cam.ac.uk, 17.04.2024

```{r}
# read decodes
decodes_tjm <-
  list.files("data/decode/tjm", pattern = "TBC", full.names = TRUE) %>%
  purrr::set_names(., basename(.)) %>%
  purrr::map(function(file) {
    readxl::read_xlsx(file, sheet = "Sheet1") %>%
      dplyr::filter(dplyr::row_number() > 15) %>%
      dplyr::transmute(
        sample_id = `Donor ID`,
        donor_id = stringr::str_sub(sample_id, end = -2),
        donor_sex = gender,
        donor_age = donor_age_at_diagnosis,
        sample_phenotype = phenotype,
        sample_tumour_status = `tumour?`,
        tube_barcode_identifier,
        Supplier_Donor_ID,
        existing_pd_id_for_donor,
        sample_ID_in_COSMIC,
        Individual
      )
  }) %>%
  dplyr::bind_rows(.id = "decode_file") %>%
  # get all other IDs in one column for a search
  tidyr::pivot_longer(cols = c(tube_barcode_identifier, Supplier_Donor_ID,
                               existing_pd_id_for_donor, sample_ID_in_COSMIC,
                               Individual),
                      names_to = "other_id_col", values_to = "other_id")
```

```{r}
# read sample sheet
tjm_supp_1 <-
  "data/decode/tjm/Samples_Tom_24-02-2021_sent.xlsx" %>%
  readxl::read_xlsx() %>%
  dplyr::transmute(
    tube_barcode_identifier = `G number`,
    sample_id_tjm = ID,
    sample_tumour_status = Region,
    donor_predisposition = Syndrome,
    donor_age = Age,
    donor_sex = `...8`
  ) 
```

```{r}
# read decode supplement 
tjm_supp_2 <-
  "data/decode/tjm/TBC_TomM_Royal_Free_Renal_Sanger_Human_Proforma_Rec20210309_WithCGP_IDS_20210315.xlsx" %>%
  readxl::read_xlsx(sheet = "Sheet2") %>%
  dplyr::transmute(
    tube_barcode_identifier = `ISBT aliquot 15Digits`,
    sample_id_tjm = ID,
    sample_type = `Derivative Type/SampleType`,
    donor_diagnosis = `Tumour type`,
    donor_predisposition = Syndrome,
  )
```

# Wrangling

1. Merge the decode files. Duplicate biopsies across multiple decode files are 
collapsed.

2. Filter out biopsies with missing BAM files. BAM files only exist for 144 / 
232 entries. According to cp19@sanger.ac.uk: "The goal of LCMing is to get the 
minimum amount of tissue needed to reach the DNA input threshold.  My cutting 
philosophy is to be right at that threshold, which means some samples will end 
up failing library prep. These proformas are all the samples I submitted for 
sequencing, not all of the samples that passed sequencing."

3. Remove donors with no normal biopsies. As the NanoSeq pipeline requires a 
at least one 'normal' sample to call somatic mutations, donors with no normal
cannot proceed.

4. Standardise `biopsy_phenotype` annotations. Different naming conventions 
were used for the same structures across the 4 decode files. I standardised
these.  

5. Standardise donor metadata. This fixes `donor_sex` categories, ensures all
rows are unique, and runs `tidyr::type_convert()`.

6. Check that `donor_id`, `sample_id`, and `biopsy_id` match up. This is to 
ensure that the IDs are consistent across the metadata levels.

```{r}
sample_sheet <-
  decodes %>%
  # 1. collapse duplicate entries across decode files
  dplyr::group_by(dplyr::across(c(-decode_file))) %>%
  dplyr::summarise(decode_file = paste(decode_file, collapse = ",")) %>%
  dplyr::ungroup() %>%
  # 2. filter out biopsies with missing bam files
  dplyr::filter(file.exists(biopsy_bam_file), biopsy_phenotype != "blank") %>%
  # 3. remove donors that have no normal biopsies
  dplyr::group_by(donor_id) %>%
  dplyr::filter("Normal" %in% sample_tumour_status) %>%
  dplyr::ungroup() %>%
  # 4. standardise sample phenotype annotations
  dplyr::mutate(
    biopsy_phenotype = dplyr::case_when(
      biopsy_phenotype %in% c("glomerulus", "glomeruls", "Glomeruli") 
        ~ "glomerulus",
      biopsy_phenotype %in% c("lymphoid", "lymphoidAggregate", "lymphTumor") 
        ~ "lymphoid_aggregate",
      biopsy_phenotype %in% c("arterial", "venous", "vessel") 
        ~ "endothelial",
      biopsy_phenotype == "tumor" 
        ~ "tumour",
      biopsy_phenotype == "hybridStromaDCT" 
        ~ "hybrid_stroma_DCT",
      TRUE ~ biopsy_phenotype)) %>%
  # 5. standardise donor metadata
  cleanup_metadata()

# 6. check that biopsy_id and sample_id and donor_id match up
check <-
  sample_sheet %>%
  dplyr::distinct(sample_id, donor_id, biopsy_id) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    donor_id_in_sample_id = stringr::str_sub(sample_id, end = -2) == donor_id,
    donor_id_in_biopsy_id = stringr::str_sub(biopsy_id, end = -9) == donor_id,
    sample_id_in_biopsy_id = stringr::str_sub(biopsy_id, end = -8) == sample_id
  ) %>%
  dplyr::filter(donor_id_in_sample_id != TRUE | is.na(donor_id_in_sample_id) |
      donor_id_in_biopsy_id != TRUE | is.na(donor_id_in_biopsy_id) |
      sample_id_in_biopsy_id != TRUE | is.na(sample_id_in_biopsy_id)
  )
if (nrow(check)) {
  message(paste("Check failed!", nrow(check), "IDs are inconsistent!"))
}
```

Now, I combine the finalised biopsy, sample, and donor metadata into a single file.

```{r}
biopsy_metadata <-
  sample_sheet %>%
  # ignore donor metadata from decodes
  dplyr::select(-c(donor_sex, donor_age)) %>%
  # add patient manifest data
  dplyr::left_join(patient_manifest)
```

## Missing ages

There are 5 donors with missing ages:

```{r}
donors_wo_ages <-
  biopsy_metadata %>%
  dplyr::filter(is.na(donor_age)) %>%
  dplyr::select(dplyr::starts_with("donor_")) %>%
  dplyr::distinct()
```

Only one donor, `PD51354`, is in the patient manifest, but does not have a listed age:

```{r}
patient_manifest %>%
  dplyr::filter(donor_id %in% donors_wo_ages$donor_id) %>%
  knitr::kable()
```

Checking the decode files, we see that none of these donors have ages listed there either.

```{r}
decodes %>%
  dplyr::bind_rows(.id = "decode_file") %>% 
  dplyr::filter(donor_id %in% donors_wo_ages$donor_id) %>%
  dplyr::distinct(donor_id, donor_age) %>%
  knitr::kable()
```

Checking the metadata from Li et al. (2022), we see that `PD44707` is listed as a 56-year-old female. 

```{r}
li2022_metadata %>%
  dplyr::filter(PD_id %in% donors_wo_ages$donor_id)
```

We can therefore amend the biopsy metadata.

```{r}
biopsy_metadata <-
  biopsy_metadata %>%
  dplyr::mutate(
    donor_age = dplyr::case_when(donor_id == "PD44707" ~ 56, TRUE ~ donor_age),
    donor_sex = dplyr::case_when(donor_id == "PD44707" ~ "F", TRUE ~ donor_sex))
donors_wo_ages <-
  donors_wo_ages %>%
  dplyr::filter(donor_id != "PD44707")
```

Checking the decode files sent by Thomas Mitchell, none of these donors have listed ages.

```{r}
decodes_tjm %>%
  dplyr::filter(donor_id %in% donors_wo_ages$donor_id |
                  other_id %in% donors_wo_ages$donor_id) %>%
  dplyr::distinct(donor_id, donor_age) %>%
  knitr::kable()
```

Checking the supplementary files sent by Thomas Mitchell, none of these donors can be found.

```{r}
codes_wo_age <-
  decodes_tjm %>%
  dplyr::filter(donor_id %in% donors_wo_ages$donor_id |
                  other_id %in% donors_wo_ages$donor_id) %>%
  {unique(c(.$donor_id, .$other_id))}
tjm_supp_1 %>%
  dplyr::filter(sample_id_tjm %in% codes_wo_age |
                  tube_barcode_identifier %in% codes_wo_age) 
```

```{r}
# view 
biopsy_metadata %>%
  head() %>%
  knitr::kable()

# write
biopsy_metadata %>%
  readr::write_tsv("out/metadata/biopsy_metadata.tsv")
```

# Generating sample sheets

## `sample_sheet.csv` for nf-kidney

The `nf-kidney` pipeline merges normal BAMs from each donor to use as a
matched normal for the `NanoSeq` pipeline. It requires a sample sheet with the
columns `donor_id`, `sample_id`, `biopsy_id`, and `sample_tumour_status`.

```{r}
sample_sheet %>%
  readr::write_csv("out/merged_normal_bams/sample_sheet.csv")
```

## `sample_sheet.csv` for NanoSeq

The `NanoSeq` pipeline requires a sample sheet with the columns `id` 
(equivalent to `biopsy_id`), `d_bam` (path to the query BAM), and `n_bam` (path
to the merged normal BAM). This sample sheet can be generated after 
`src/01_merge_normal_bams.sh` has successfully run the `nf-kidney` pipeline. 

```{r}
# transmute
sample_sheet_nanoseq <-
  sample_sheet %>%
  dplyr::transmute(
    id = biopsy_id,
    d_bam = biopsy_bam_file,
    n_bam = file.path(getwd(),
                      "out/merged_normal_bams",
                      paste0(donor_id, "_merged.bam"))) %>%
  # check files exist
  dplyr::filter(file.exists(d_bam) & file.exists(n_bam))

# write
sample_sheet_nanoseq %>%
  readr::write_csv("out/nanoseq/sample_sheet.csv")

# view
sample_sheet_nanoseq %>%
  head() %>%
  knitr::kable()
```

# Outputs

There are three key outputs from this script:

1. A sample sheet used as input to the `nf-kidney` pipeline (see `./nf-kidney/`)

2. A sample sheet used as input to the `NanoSeq` pipeline (see [cancerit/NanoSeq](https://github.com/cancerit/NanoSeq))

3. A harmonised metadata file that contains clean donor and sample information for the project

