---
title: "03_visualise"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include = F, message = F, warning = F, class.source = 'fold-hide'}
# Rscript -e "rmarkdown::render('src/03_visualise.Rmd', output_dir = 'out/nanoseq/analysis/')" # nolint: line_length_linter.

# knitr opts
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(warning = FALSE, dpi = 300, message = FALSE)

# dirs
nanoseq_dir <- "out/nanoseq/outNextflow/"
dir.create("out/nanoseq/analysis/", showWarnings = FALSE)

# libraries
library(magrittr)
library(ggplot2)

# functions ####

# function: get files in nanoseq dir from prefix
get_files <- function(biopsy_ids, ext) {
  files <-
    biopsy_ids %>%
    purrr::set_names(., .) %>%
    purrr::map(function(biopsy_id) {
      list.files(file.path(nanoseq_dir, biopsy_id, "post"),
                 full.names = TRUE, pattern = ext)
    }) %>%
    # remove empty elements
    Filter(length, .)

  if (grepl("tsv", ext) == TRUE) {
    # load tsv
    out <-
      files %>%
      purrr::map_df(function(file) {
        file %>% read.table() %>% tibble::as_tibble(rownames = "row")
      }, .id = "biopsy_id")
  } else if (grepl("vcf.gz", ext) == TRUE) {
    # load vcf
    out <-
      files %>%
      purrr::map_df(function(file) {
        vcfR::read.vcfR(file, verbose = FALSE)@fix %>%
          tibble::as_tibble() %>%
          # only get variants that pass
          dplyr::filter(FILTER == "PASS") %>%
          # count variants per biopsy
          dplyr::mutate(n = dplyr::n())
      }, .id = "biopsy_id")
  } else {
    stop("Extension not recognised")
  }

  # return
  return(out)
}

# function: plot trinuc distributions
plot_trinucs <- function(trinucs) {
  trinucs %>%
    ggplot(aes(x = trinuc, y = trint_subst_obs, fill = facet)) +
    geom_col(width = 0.5) +
    scale_fill_manual(values = colours_vec) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, family = "courier",
                                     size = 5),
          legend.position = "none",
          panel.grid = element_blank()) +
    labs(x = "substitution", y = "observed mutation counts")
}

# colours and orders ####

# set order/colours of biopsy
# "The correct pathway within the nephron starts with the glomerulus followed
# by the Bowman's capsule, proximal convoluted tubule, the descending loop of
# Henle, the ascending loop of Henle, distal convoluted tubule, and collecting
# duct. The glomerulus is found within the Bowman's capsule.
phenotype_order <-
  c("glomerulus" = "#66A61E",
    "PCT" = "#E6AB02",
    "DCT" = "#D95F02",
    "lymphoid_aggregate" = "#7570B3",
    "endothelial" = "#E7298A",
    "tumour" = "black")

# set order/colours of trinucleotide subs
colours_vec <-
  c("C>A" = "deepskyblue",
    "C>G" = "black",
    "C>T" = "firebrick2",
    "T>A" = "gray",
    "T>C" = "darkolivegreen3",
    "T>G" = "rosybrown2")
```

# Introduction

This script analyses the output from the `NanoSeq` pipeline for the project 
**NanoSeq - (CP) Normal and cancer-predisposed kidney (2731)**.

# Load metadata

```{r load_metadata}
# load sample sheet
sample_sheet <-
  readr::read_csv("out/nanoseq/sample_sheet.csv")

# load biopsy metadata and reorder phenotypes
biopsy_metadata <-
  readr::read_tsv("out/metadata/biopsy_metadata.tsv") %>%
  dplyr::mutate(biopsy_phenotype = factor(biopsy_phenotype,
                                          levels = names(phenotype_order)))
```

# NanoSeq outputs

The output files from the NanoSeq pipeline are located in the `post/` directory.

```{bash}
ls -d out/nanoseq/outNextflow/*/post/ | head -1 | xargs ls
```

According to the [NanoSeq GitHub](https://github.com/cancerit/NanoSeq), these
are the most relevant output files from the pipeline:

* `muts.vcf.gz / muts.tsv`: substitutions called in vcf and tsv format. "PASS" 
substitutions are those not filtered by the common SNP and noisy sites masks 
(see Genomic masks).

* `indels.vcf.gz`: indel calls

* `burden.masked-vs-unmasked.pdf`: estimated burden before and after filtering 
common SNPs. Provides a qualitative view on contamination.

* `mut_burden.tsv`: estimated substitution burden with Poisson confidence
intervals. The corrected burden shows the burden after normalizing observed 
trinucleotide frequencies to the genomic trinucleotide frequencies.

* `trinuc-profiles.pdf` / `trint_subs_obs_corrected.tsv` / 
`trint_counts_and_ratio2genome.tsv`: Trinucleotide substitution profiles 
(observed and corrected), using the trinucleotide substitution counts in 
`trint_subs_obs_corrected.tsv` and the normalization of trinucleotide 
frequencies in `trint_counts_and_ratio2genome.tsv`. Normalization is required 
because NanoSeq results are depleted of trinucleotides overlapping the 
restriction site and of CpGs due to extensive filtering of common SNPs.

* `cov.bed.gz`: large file containing the effective duplex coverage for each 
genomic site, also showing the trinucleotide context of each site. This file is 
required to to calculate burdens and substitution profiles in sets of specific 
genomic regions (e.g. highly expressed genes, heterochromatin, ...).

* `subst_asym.pdf` / `subst_asym_and_rates_binned.pdf` / `subst_asym_binned.tsv` 
/ `subst_asym.pvals` / `subst_asym.tsv`: These files are not generally needed 
for NanoSeq analysis. They were originally used to detect asymmetries in the 
original DuplexSeq & BotSeqS protocols.

* `mismatches.trinuc-profile.pdf` / `mismatches.subst_asym.pdf` / 
`mismatches.subst_asym.pvals` / `mismatches.subst_asym.tsv`: These files show 
the asymmetries and pyrimidine/purine-based trinucleotide substitution profiles 
for single-strand consensus calls. These profiles are useful to understand DNA 
damage during library preparation.

* `DSC_errors_per_channel.pdf` / `DSC_estimated_error_rates.pdf` / 
`estimated_error_rates.tsv` / `SSC-mismatches-Both.triprofiles.tsv` / 
`SSC-mismatches-Purine.triprofiles.tsv` / 
`SSC-mismatches-Pyrimidine.triprofiles.tsv`: Based on the independent error 
rates in the purine and pyrimidine channels (e.g. G>T and C>A), we calculate the 
probability of having independent errors affecting both strands and resulting in 
double-strand consensus.

# Load the NanoSeq outputs

I want to visualise the counts and burdens of indels and SNVs, and to plot the
distribution of trinucleotide substitutions. First, we load the relevant files.

```{r load_post}
post <-
  list("mut_burden" = ".mut_burden.tsv$",
       "trint_subs_obs_corrected" = ".trint_subs_obs_corrected.tsv$",
       "indel" = ".indel.vcf.gz$",
       "muts" = ".muts.vcf.gz$") %>%
  purrr::map(~ get_files(sample_sheet$id, .x))
```

I also want to test for contamination. I load the `verifyBAMid.txt` files for
these biopsies.

```{r verify_bam_id}
# TODO: implement this
verify <-
  post %>%
  purrr::map(~ .$biopsy_id) %>%
  unlist() %>%
  unique() %>%
  purrr::map(function(biopsy_id) {
    paste0(nanoseq_dir, biopsy_id, "/verifyBAMid/", biopsy_id,
           "_duplex.verifyBAMid.txt") %>%
      readr::read_delim(delim = ":", skip = 3)
  })
```

Next, we combine the burdens from SNVs and from indels into a single long-form
dataframe.

* The `variant_type` column is either `snv` or `indel`. 

* The `value_type` column denotes what the `value` column is measuring. `count` 
is the number of each mutation type, `total` is the number of sites sequenced, 
and `burden` is the proportion of all sites sequenced that are of that mutation 
type (i.e. `count` / `total`). 

N.B. `total` values for indels and SNVs will be the same in each biopsy. 

```{r get_burdens}
# get all burdens
burdens <-
  post[["mut_burden"]] %>%
  # standardise names
  dplyr::transmute(biopsy_id, snv_count = muts, snv_burden = burden,
                   total, snv_total = total, indel_total = total) %>%
  dplyr::left_join(post[["indel"]] %>%
                     dplyr::select(biopsy_id, indel_count = n) %>%
                     dplyr::distinct()) %>%
  # calculate indel burden
  dplyr::mutate(indel_burden = indel_count / total) %>%
  # convert to long format
  tidyr::pivot_longer(cols = c(dplyr::ends_with("_count"),
                               dplyr::ends_with("_burden"),
                               dplyr::ends_with("_total")),
                      names_pattern = "([^.]+)\\_([^.]+)$",
                      names_to = c("variant_type", "value_type")) %>%
  # add metadata
  dplyr::left_join(biopsy_metadata)

# view sample
burdens %>%
  head() %>%
  knitr::kable()
```

# Visualise

## Burdens, counts, and totals

### Burdens, counts, and totals vs biopsy ID

```{r plot_burdens_vs_biopsy, class.source = 'fold-hide', fig.width = 12, fig.height = 12}
# TODO: reorder_within by age
burdens %>%
  ggplot(aes(x = biopsy_id, y = value, fill = biopsy_phenotype)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  scale_fill_manual(values = phenotype_order) +
  ggh4x::facet_nested(variant_type + value_type ~ biopsy_phenotype,
                      scales = "free", space = "free_x") +
  theme_bw() +
  theme(panel.spacing = unit(0.1, "cm", data = NULL))
```

### Burdens, counts, and totals vs donor age

```{r plot_burden_vs_age, class.source = 'fold-hide'}
burdens %>%
  dplyr::filter(!is.na(donor_age)) %>%
  ggplot(aes(x = donor_age, y = value, colour = biopsy_phenotype)) +
  geom_point() +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  scale_colour_manual(values = phenotype_order) +
  ggh4x::facet_nested(value_type ~ variant_type, scales = "free",
                      space = "free_x") +
  theme_bw()
```

### Median burden per donor vs donor age

```{r plot_med_burden_vs_age, class.source = 'fold-hide'}
burdens %>%
  dplyr::filter(!is.na(donor_age)) %>%
  dplyr::group_by(donor_id) %>%
  dplyr::mutate(median_value = median(value),
                n_biopsies = dplyr::n_distinct(biopsy_id)) %>%
  ggplot(aes(x = donor_age, y = median_value, colour = donor_predisposition,
             size = n_biopsies)) +
  geom_point() +
  theme_bw()
```

## Trinucleotide substitutions

```{r trinucs_p_dat}
# wrangle trinucleotide substitutions
p_dat <-
  post[["trint_subs_obs_corrected"]] %>%
  dplyr::rename(substitution = row) %>%
  # add metadata
  dplyr::left_join(biopsy_metadata) %>%
  dplyr::mutate(facet = paste0(stringr::str_sub(substitution, 2, 2),
                               ">",
                               stringr::str_sub(substitution, 5, 5)),
                trinuc = stringr::str_sub(substitution, 1, 3))
```

### By donor

Trinucleotide substitutions are plotted for all biopsies of each donor. Counts 
in each biopsy are summed per donor.

```{r plot_trinucs_by_donor, class.source = 'fold-hide', fig.height = 2.5, fid.width = 10}
p_dat %>%
  {split(., .$donor_id)} %>%
  purrr::map(function(df) {
    title <- paste0(unique(df$donor_id),
                    "\n", dplyr::n_distinct(df$biopsy_id), " biopsies, ",
                    sum(df$trint_subst_obs), " substitutions")
    df %>%
      plot_trinucs() +
      facet_grid(~ facet, scales = "free_x") +
      ggtitle(title)
  })
```

### By biopsy phenotype

Trinucleotide substitutions are plotted for all patients for each biopsy
phenotype. Counts in each patient are summed per biopsy phenotype.

```{r plot_trinuc_by_biopsy_phenotype, class.source = 'fold-hide', fig.height = 2.5, fid.width = 10}
p_dat %>%
  {split(., .$biopsy_phenotype)} %>%
  purrr::walk(function(df) {
    title <- paste0(unique(df$biopsy_phenotype),
                    "\n", dplyr::n_distinct(df$biopsy_id), " biopsies, ",
                    sum(df$trint_subst_obs), " substitutions")
    df %>%
      plot_trinucs() +
      facet_grid(~ facet, scales = "free_x") +
      ggtitle(title)
  })
```

### By donor x biopsy phenotype

Trinucleotide susbtitutions are plotted for 

```{r plot_trinuc_by_donor_x_phenotype, class.source = 'fold-hide', fig.height = 2.5, fid.width = 10}
p_dat %>%
  {split(., .$donor_id)} %>%
  purrr::map(function(df) {
    df %>%
      {split(., as.character(.$biopsy_phenotype))} %>%
      purrr::map(function(df2) {
        title <- paste0(unique(df2$donor_id), " ", unique(df2$biopsy_phenotype),
                        "\n", dplyr::n_distinct(df2$biopsy_id), " biopsies, ",
                        sum(df2$trint_subst_obs), " substitutions")
        df2 %>%
          plot_trinucs() +
          theme(strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0)) +
          facet_grid(~ facet, scales = "free_x") +
          ggtitle(title)
      })
  })
```